---
title: "Data Cleaning"
author: "Shonushka Sawant"
format: 
  html:
    self-contained: true
---

```{r}
library(dplyr)
library(tidyr)
library(rstatix)
library(ggplot2)
```

```{r}
mmr_cov <- read.csv("USA/data/all/measles-USA-by-mmr-coverage.csv")
head(mmr_cov)
```


```{r}
drop_na.tbl_ts <- function(ts)  tsibble::as_tsibble(tidyr:::drop_na.data.frame(ts))
```

```{r}

#drop rows with NA for vaccine percentage
mmr_cov <- mmr_cov %>% drop_na(estimate_pct)

#create a column with percent cast to numeric
mmr_cov$num_pct <- mmr_cov$estimate_pct

mmr_cov$num_pct <- as.numeric(sub("%","",mmr_cov$num_pct))/100
```

```{r}
#summary statistics for vaccine coverage, per state
group_by(mmr_cov, geography) %>% 
  get_summary_stats(num_pct)
```

```{r}
head(mmr_cov)
```
```{r}
#summary statistics for vaccine coverage, per year
summ_year <- group_by(mmr_cov, school_year) %>% 
  get_summary_stats(num_pct)

summ_year
```

```{r}
#plot median vaccination rate by year

summ_year %>%
  tail(10) %>%
  ggplot( aes(x=school_year, y=median)) +
    geom_line(color="black") +
    geom_point()
```

Median vaccination rates among states in the USA appear to be showing a downward trend starting from the 2019-20200 school year. We will now look at the trajectories for the states with the lowest vaccination rates in 2023-2024.

```{r}
mmr_cov2023 <- mmr_cov %>% filter(school_year == "2023-24")
head(mmr_cov2023 %>% arrange(num_pct))
```

The states with the five lowest vaccination rates are Idaho (79.6%), Alaska (84.3%), Wisconsin (84.8%), Minnesota (87.0%), and Florida (88.1%).

Compare these rates to the 2009-10 school year:

```{r}
low_2023 <- c("Idaho", "Alaska", "Wisconsin", "Minnesota", "Florida")
mmr_cov2009 <- mmr_cov %>% filter(geography %in% low_2023, school_year == "2009-10")
mmr_cov2009 %>% arrange(num_pct)
```

Vaccination rates for these states were notably higher in 2009-10 (excepting Alaska, which did not report its vaccination rate that year.)

```{r}

mmr_cov_low <- mmr_cov %>% filter(geography %in% low_2023)
ggplot(mmr_cov_low, aes(geography, num_pct, fill = geography)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  guides(fill = "none") +
  labs(x = "", y = "Vaccination rate")
```

Wisconsin and Alaska show the greatest variation in vaccination rates, of the five states with the lowest rates in 2023-24.

```{r}
#at this point we will add a numeric column for the dates.
mmr_cov$num_years <- substr(mmr_cov$school_year, 1, 4)
mmr_cov$num_years <- as.numeric(mmr_cov$num_years)

head(mmr_cov)
```

```{r}
#fitting linear mixed effects model.
#this is inaccurate, just for practice with lmer for now.
library(lme4)
lin_0 <- lmer(num_years ~ 1 + (1 | geography), data = mmr_cov)
summary(lin_0)
```

```{r}
library(Epi)
ci.lin(lin_0)
```

